
---
title: "German Credit Case Study"
author: "Gbadamosi Muibat Asabi"
date: "`r Sys.Date()`"
---

# Problem statement:

## Context: 
A decision rule for loan approval is necessary to avoid losses for banks. The goal is to create a prediction model that will help the bank determine whether to offer a loan to a prospective applicant based on their demographic and socioeconomic profiles.Each item in the dataset represents a person who accepts credit from a bank and is rated as a good(1) or bad(0) credit risk based on the characteristics.

## Objective
The objective is to create a prediction model that can identify potential loan applications as excellent or bad credit risks.

# Libraries Used
The following are the libraries used in the datasets
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(caret)
library(readr)
library(rmarkdown)
library(corrplot)
library(caTools)

```

# Loading the dataset
```{r}
credit_data <- read_csv("C:/Users/user/Documents/German_Credit (1).csv")

```
# Dataset Description
The dataset contains the following attributes: They are four(4) Numeric variables and six(6) character variables. Totally,they are 1000 columns and 10 rows

- **Age**: Age in years (Numeric)

- **Sex**: Gender of the applicant (Categories: male, female)

- **Job**: Job status (Categories: unskilled and non-resident, unskilled and resident, skilled, highly skilled)

- **Housing**: Housing status (Categories: own, rent, free)

- **Saving accounts**: Status of savings accounts (Categories: little, moderate, quite rich, rich)

- **Checking account**: Status of checking accounts (Categories: little, moderate, rich)

- **Credit amount**: Amount of credit in Deutsche Mark (Numeric)

- **Duration**: Duration for which the credit is given in months (Numeric)

- **Purpose**: Purpose of the credit (Categories: car, furniture/equipment, radio/TV, domestic appliances, repairs, education, business, vacation/others)

- **Risk**: Credit risk (0 - Not at risk, 1 - At risk/defaulter)

# Structure of the data
```{r}
str(credit_data)
```

#Summary of the Dataset
```{r}
summary(credit_data)
```




# Check for missing values
No missing value was found.
```{r}
#Check for missing values
sum(is.na(credit_data))
```





# EDA(Exploratory Data Analysis)
The EDA is to explore the distribution of categorical and Continuous variables in the dataset and their relationship with the target variable `Risk`.

# Categorial Variables
The categorical variables includes the Sex,Job,Housing,Savingaccount,Checkingaccount,Purpose.

## Sex distribution

```{r}
# Count and proportion of each category in Sex
table(credit_data$Sex)
prop.table(table(credit_data$Sex))
```

```{r}
# Barplot for Sex
ggplot(credit_data, aes(x=Sex)) +
  geom_bar(fill="lightgreen") +
  labs(title="Distribution of Sex", x="Sex", y="Count") +
  theme_minimal()
```

```{r}
# Count plot of Sex vs Risk
ggplot(credit_data, aes(x = Sex, fill = Risk)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Risk by Sex", x = "Sex", y = "Count") +
  scale_fill_manual(values = c("blue", "red"))
```

```{r}
# Chi-square test for Sex and Risk
chisq_test_sex <- chisq.test(table(credit_data$Sex, credit_data$Risk))
print(chisq_test_sex)
```

## Job distribution
```{r}
# Count and proportion of each category in Job
table(credit_data$Job)
prop.table(table(credit_data$Job))
```

```{r}
# Barplot for Job
ggplot(credit_data, aes(x=Job)) +
  geom_bar(fill="lightgreen") +
  labs(title="Distribution of Job", x="Job", y="Count") +
  theme_minimal()

```

```{r}
# Barplot for Job
ggplot(credit_data, aes(x = Job, fill = Risk)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Risk by Job", x = "Job", y = "Count") +
  scale_fill_manual(values = c("blue", "red"))

```

## Housing Distribution

```{r}
# Count and proportion of each category in Housing
table(credit_data$Housing)
prop.table(table(credit_data$Housing))
```

```{r}
# Barplot for Housing

ggplot(credit_data, aes(x=Housing)) +
  geom_bar(fill="lightcoral") +
  labs(title="Distribution of Housing", x="Housing", y="Count") +
  theme_minimal()
```

```{r}
# Count plot of Housing vs Risk
ggplot(credit_data, aes(x = Housing, fill = Risk)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Risk by Housing", x = "Housing", y = "Count") +
  scale_fill_manual(values = c("blue", "red"))
```


##Saving Distribution

```{r}
# Count and proportion of each category in Housing
table(credit_data$`Saving accounts`)
prop.table(table(credit_data$`Saving accounts`))
```

```{r}
# Barplot for Housing

ggplot(credit_data, aes(x=`Saving accounts`)) +
  geom_bar(fill="lightcoral") +
  labs(title="Distribution of Saving account", x="Saving account", y="Count") +
  theme_minimal()
```

```{r}
# Count plot of Saving Accounts vs Risk
ggplot(credit_data, aes(x =`Saving accounts`, fill = Risk)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Risk by Saving Accounts", x = "Saving Accounts", y = "Count") +
  scale_fill_manual(values = c("blue", "red"))
```

## Checking account Distribution

```{r}
# Count and proportion of each category in Checking account
table(credit_data$`Checking account`)
prop.table(table(credit_data$`Checking account`))
```

```{r}
# Barplot for checking account

ggplot(credit_data, aes(x=`Checking account`)) +
  geom_bar(fill="lightcoral") +
  labs(title="Distribution of Checking account", x="Checking account", y="Count") +
  theme_minimal()
```

```{r}
# Count plot of Checking Account vs Risk
ggplot(credit_data, aes(x = `Checking account`, fill = Risk)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Risk by Checking Account", x = "Checking Account", y = "Count") +
  scale_fill_manual(values = c("blue", "red"))
```

##Purpose Distribution

```{r}
# Count and proportion of each category in Purpose
table(credit_data$Purpose)
prop.table(table(credit_data$Purpose))

```

```{r}
# Barplot for Purpose

ggplot(credit_data, aes(x=Purpose)) +
  geom_bar(fill="lightcoral") +
  labs(title="Distribution of Purpose", x="Purpose", y="Count") +
  theme_minimal()

```

```{r}
# Count plot of Purpose vs Risk
ggplot(credit_data, aes(x = Purpose, fill = Risk)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Risk by Purpose", x = "Purpose", y = "Count") +
  scale_fill_manual(values = c("blue", "red")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
# Continuous Variables
The continuous variables includes the Age,Duration,Creditamount
## Age Distribution
```{r}
# Histogram of Age
ggplot(credit_data, aes(x=Age)) +
  geom_histogram(binwidth=5, fill="steelblue", color="black") +
  labs(title="Distribution of Age", x="Age (years)", y="Count") +
  theme_minimal()
```

```{r}
# Boxplot of Age by Risk
ggplot(credit_data, aes(x=Risk, y=Age, fill=Risk)) +
  geom_boxplot() +
  labs(title="Age vs Risk", x="Risk", y="Age (years)") +
  theme_minimal()

```

##  Credit Amount Distribution
```{r}
# Histogram of Credit Amount
ggplot(credit_data, aes(x=`Credit amount`)) +
  geom_histogram(binwidth=1000, fill="lightgreen", color="black") +
  labs(title="Distribution of Credit Amount", x="Credit Amount (DM)", y="Count") +
  theme_minimal()
```


```{r}
# Boxplot of Credit Amount by Risk
ggplot(credit_data, aes(x=Risk, y=`Credit amount`, fill=Risk)) +
  geom_boxplot() +
  labs(title="Credit amount vs Risk", x="Risk", y="Credit amount(years)") +
  theme_minimal()

```
##  Duration Distribution

This shows that people at risk of defaulting (1) generally have longer credit durations than those not at risk (0). The median duration for defaulters is around 30 months, while for non-defaulters, it's 18-20 months. Therefore the longer credit durations,the higher risk of default.
```{r}
# Histogram of Duration
ggplot(credit_data, aes(x=Duration)) +
  geom_histogram(binwidth=6, fill="coral", color="black") +
  labs(title="Distribution of Duration", x="Duration (months)", y="Count") +
  theme_minimal()

```

```{r}
# Boxplot of Duration by Risk
ggplot(credit_data, aes(x=Risk, y=Duration, fill=Risk)) +
  geom_boxplot() +
  labs(title="Duration vs Risk", x="Risk", y="Duration (months)") +
  theme_minimal()

```
# Heatmap of continuous variables
Heatmaps are typically used for numerical data to visualize correlations or intensities.The figures shows that there is a moderate correlation at 0.62, indicating a stronger positive relationship. As the Credit amount increases, the Duration tends to increase as well. while Age has little to no correlation with the other two variables.
```{r}
# Correlation matrix
continuous_vars <-credit_data[ c("Age", "Credit amount", "Duration")]
cor_matrix <- cor(continuous_vars)
corrplot(cor_matrix, method = "color", addCoef.col = "black", 
         tl.col = "black", number.cex = 0.8, cl.pos = "r", 
         col = colorRampPalette(c("red", "white", "blue"))(200),
         title = "Heatmap of Correlation for Continuous Variables", 
         mar = c(0, 0, 1, 0))

```

# Encoding Categorical Variables
Many machine learning algorithms require numeric inputs. Categorical variables like Sex,Job,Housing,Savingaccount,Checkingaccount,Purpose  need to be encoded into numerical format using Label encoding.

```{r}
# Convert categorical variables to factors first if they are not already
credit_data$Sex <- as.factor(credit_data$Sex)
credit_data$Job <- as.factor(credit_data$Job)
credit_data$Housing <- as.factor(credit_data$Housing)
credit_data$Saving.accounts <- as.factor(credit_data$`Saving accounts`)
credit_data$Checking.account <- as.factor(credit_data$`Checking account`)
credit_data$Purpose <- as.factor(credit_data$Purpose)
credit_data$Risk <- as.factor(credit_data$Risk)



# Label encode the factors (convert to integers)
credit_data$Sex_encoded <- as.numeric(credit_data$Sex)
credit_data$Job_encoded <- as.numeric(credit_data$Job)
credit_data$Housing_encoded <- as.numeric(credit_data$Housing)
credit_data$Saving_accounts_encoded <- as.numeric(credit_data$`Saving accounts`)
credit_data$Checking_account_encoded <- as.numeric(credit_data$`Checking account`)
credit_data$Purpose_encoded <- as.numeric(credit_data$Purpose)
credit_data$Risk_encoded <- as.numeric(credit_data$Risk)

```

```{r}
str(credit_data)

```


# Standardizing/Normalizing Continuous Variables
For models like Logistic Regression and Neural Networks, it's important to standardize or normalize continuous variables like Age, Credit.amount, and Duration.

```{r}
# Scaling numeric variables (standardization)
credit_data$Age <- scale(credit_data$Age)
credit_data$Credit.amount <- scale(credit_data$`Credit amount` )
credit_data$Duration <- scale(credit_data$Duration)

```

```{r}
str(credit_data)

```

# Standardizing/Normalizing Continuous Variables
For models like Logistic Regression and Neural Networks, it's important to standardize or normalize continuous variables like Age, Credit.amount, and Duration.


```{r}
# Scaling numeric variables (standardization)
credit_data$Age <- scale(credit_data$Age)
credit_data$Credit.amount <- scale(credit_data$`Credit amount` )
credit_data$Duration <- scale(credit_data$Duration)
```
# Traning and Testing datasets

The datasets is splited into Training and Testing in the ratio of 90 - 10.

```{r}
# spliting the data
set.seed(234)
in.train <- createDataPartition(as.factor(credit_data$Risk), p=0.8, list=FALSE)
german_credit.train <- credit_data[in.train,]
german_credit.test <-credit_data[-in.train,]

```

```{r}
# train data
nrow(train_data)
```

```{r}
# test data
nrow(test_data)
```

# Model Choices

There are several models for binary classification(i.e risk as 0 or 1) but i will be using Logistics regression and random forest

#Logistic Regression Model

```{r}
# Check the unique values in the Risk variable
unique(train_data$Risk)


# If Risk is a factor with levels "Good" and "Bad", convert it to 0 and 1
train_data$Risk <- as.factor(train_data$Risk)
test_data$Risk <- as.factor(test_data$Risk)


# You may need to convert it to binary numeric values (0 or 1) for logistic regression
train_data$Risk <- ifelse(train_data$Risk == "Good", 0, 1)
test_data$Risk <- ifelse(test_data$Risk == "Good", 0, 1)

```

```{r}
#  Checking levels of factors in the data
lapply(german_credit.train, function(x) if(is.factor(x)) levels(x))

#  Identifying  columns with only one level
single_level_vars <- sapply(german_credit.train, function(x) length(unique(x)) == 1)

#  Removing  columns with only one unique value
german_credit.train.clean <- german_credit.train[, !single_level_vars]

#  Refiting  the model with the cleaned data
credit.glm0 <- glm(Risk ~ ., family = binomial, data = german_credit.train.clean)

```

```{r}
# Summary of the model
summary(credit.glm0)
```

```{r}
# Make predictions on the test set
predicted_prob <- predict(credit.glm0, newdata = german_credit.test, type = "response")
# Convert probabilities to class labels (assuming 0.5 threshold for binary classification)
predicted_class <- ifelse(predicted_prob > 0.5, 1, 0)
# Confusion matrix using table function
actual_class <- german_credit.test$Risk
confusion_matrix <- table(Predicted = predicted_class, Actual = actual_class)
print(confusion_matrix)


```

```{r}

```
